C:
int MPI_Accumulate
   (const void *origin_addr, int origin_count,MPI_Datatype origin_datatype,
    int target_rank,MPI_Aint target_disp, int target_count,MPI_Datatype target_datatype,
    MPI_Op op, MPI_Win win)
int MPI_Raccumulate
   (const void *origin_addr, int origin_count,MPI_Datatype origin_datatype, 
    int target_rank,MPI_Aint target_disp, int target_count,MPI_Datatype target_datatype, 
    MPI_Op op, MPI_Win win,MPI_Request *request)

Input Parameters

origin_addr : Initial address of buffer (choice).
origin_count : Number of entries in buffer (nonnegative integer).
origin_datatype : Data type of each buffer entry (handle).
target_rank : Rank of target (nonnegative integer).
target_disp : Displacement from start of window to beginning of target buffer (nonnegative integer).
target_count : Number of entries in target buffer (nonnegative integer).
target_datatype : Data type of each entry in target buffer (handle).
op : Reduce operation (handle).
win : Window object (handle).

Output Parameter

MPI_Raccumulate: RMA request
IERROR (Fortran only): Error status (integer).

Fortran:

MPI_ACCUMULATE
   (ORIGIN_ADDR, ORIGIN_COUNT, ORIGIN_DATATYPE,
    TARGET_RANK,TARGET_DISP, TARGET_COUNT, TARGET_DATATYPE, 
    OP, WIN, IERROR)
<type> ORIGIN_ADDR(*)
INTEGER(KIND=MPI_ADDRESS_KIND) :: TARGET_DISP
INTEGER :: ORIGIN_COUNT, ORIGIN_DATATYPE, 
           TARGET_RANK, TARGET_COUNT,TARGET_DATATYPE, 
           OP, WIN, IERROR
MPI_RACCUMULATE
   (ORIGIN_ADDR, ORIGIN_COUNT, ORIGIN_DATATYPE,
    TARGET_RANK,TARGET_DISP, TARGET_COUNT, TARGET_DATATYPE, 
    OP, WIN, REQUEST, IERROR)
<type> ORIGIN_ADDR(*)
INTEGER(KIND=MPI_ADDRESS_KIND) :: TARGET_DISP
INTEGER :: ORIGIN_COUNT, ORIGIN_DATATYPE, TARGET_RANK, TARGET_COUNT, TARGET_DATATYPE,
           OP, WIN, REQUEST, IERROR

Python:
MPI.Win.Accumulate(self, origin, int target_rank, target=None, Op op=SUM)
