% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This text file is part of the source of 
%%%% `Parallel Programming in MPI and OpenMP'
%%%% by Victor Eijkhout, copyright 2012-2021
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A communicator is an object describing a group of processes. In many 
applications all processes work together closely coupled, and the
only communicator you need is \indexmpishow{MPI_COMM_WORLD}, the
group describing all processes that your job starts with.

In this chapter you will see ways to make new groups of MPI processes:
subgroups of the original world communicator.
Chapter~\ref{ch:mpi-proc} discusses dynamic process management, which, while
not extending \lstinline{MPI_COMM_WORLD} does extend the set of
available processes.
That chapter also discusses the `sessions model',
which is another way to constructing communicators.

\Level 0 {Basic communicators}
\label{sec:comm-basic}

There are three predefined communicators:
\begin{itemize}
\item \indexmpishow{MPI_COMM_WORLD} comprises all processes that were started 
  together by \indextermtt{mpiexec} (or some related program).
\item \indexmpishow{MPI_COMM_SELF} is the communicator that contains only
   the current process.
 \item \indexmpishow{MPI_COMM_NULL} is the invalid communicator.
   This values results
   \begin{itemize}
   \item when a communicator is freed; see section~\ref{sec:communicators};
   \item as error return value from routines that construct communicators;
   \item for processes outside a created Cartesian communicator
     (section~\ref{sec:cart-comm-create});
   \item on non-spawned processes when querying their parent
     (section~\ref{sec:mpi-inter-query}).
   \end{itemize}

\end{itemize}
These values are constants, though not necessarily compile-time constants.
Thus, they can not be used in switch statements, array declarations,
or \n{constexpr} evaluations.

If you don't want to write \indexmpishow{MPI_COMM_WORLD} repeatedly, you can
assign that value to a variable of type \indexmpidef{MPI_Comm}.

Examples:
\lstset{language=C}
\begin{lstlisting}
// C:
#include <mpi.h>
MPI_Comm comm = MPI_COMM_WORLD;
\end{lstlisting}

\lstset{language=Fortran}
\begin{lstlisting}
!! Fortran 2008 interface
use mpi_f08
Type(MPI_Comm) :: comm = MPI_COMM_WORLD
\end{lstlisting}

\begin{lstlisting}
!! Fortran legacy interface
#include <mpif.h>
Integer :: comm = MPI_COMM_WORLD
\end{lstlisting}
\lstset{language=C}

\begin{pythonnote}{Communicator types}
\lstset{language=python}
\begin{lstlisting}
comm = MPI.COMM_WORLD
\end{lstlisting}
\end{pythonnote}

\begin{mplnote}{Predefined communicators}
  The \indexmplshow{environment} namespace has the equivalents
  of \indexmpishow{MPI_COMM_WORLD} and \indexmpishow{MPI_COMM_SELF}:
\begin{lstlisting}
const communicator& mpl::environment::comm_world();
const communicator& mpl::environment::comm_self();    
\end{lstlisting}
There doesn't seem to be an equivalent of \indexmpishow{MPI_COMM_NULL}.
\end{mplnote}

\begin{mplnote}{Raw communicators}
  Should you need the \indexmpishow{MPI_Comm} object
  contained in an MPL \indexmplshow{communicator},
  there is an access function \indexmplshow{native_handle}.
  \csnippetwithoutput{mplcommraw}{examples/mpi/mpl}{rawcompare}
\end{mplnote}

You can name your communicators with \indexmpidef{MPI_Comm_set_name}, which
could improve the quality of error messages when they arise.

\Level 0 {Duplicating communicators}
\label{sec:comm-dup}

\input chapters/mpi-commdup

\Level 0 {Sub-communicators}
\label{sec:communicators}
\index{communicator|(}

In many scenarios you divide a large job over all the available processors.
However, your job may have two or more parts that can be considered as
jobs by themselves. In that case it makes sense to divide your processors
into subgroups accordingly.

Suppose for instance that you are running a simulation where inputs are generated,
a~computation is performed on them, and the results of this computation
are analyzed or rendered graphically. You could then consider dividing your
processors in three groups corresponding to generation, computation, rendering.
%
As long as you only do sends and receives, this division works fine. However,
if one group of processes needs to perform a collective operation, you don't
want the other groups involved in this. Thus, you really want the three groups
to be distinct from each other.

In order to make such subsets of processes, MPI has the mechanism of
taking a subset of \indexmpishow{MPI_COMM_WORLD}
(or other communicator)
and turning that subset into a new communicator.

Now you understand why the MPI collective calls had an argument for the
communicator: a~collective involves all processes of that communicator.
By making a communicator that contains a subset of all available processes,
you can do a collective on that subset.

The usage is as follows:
\begin{itemize}
\item You create a new communicator with routines such as
  \indexmpishow{MPI_Comm_dup} (section~\ref{sec:comm-dup}),
  \indexmpishow{MPI_Comm_split} (section~\ref{sec:comm-split}),
  \indexmpishow{MPI_Comm_create} (section~\ref{sec:mpi-comm-group}),
  \indexmpishow{MPI_Intercomm_create} (section~\ref{sec:mpi-intercomm}),
  \indexmpishow{MPI_Comm_spawn} (section~\ref{sec:mpi-dynamic});
\item you use that communiator for a while;
\item and you call \indexmpidef{MPI_Comm_free} when you are done with it;
  this also sets the communicator variable to \indexmpishow{MPI_COMM_NULL}.
  A~similar routine, \indexmpidef{MPI_Comm_disconnect} waits for all pending
  communication to finish. Both are collective.
\end{itemize}

\Level 1 {Scenario: distributed linear algebra}

For \emph{scalability} reasons, matrices should often be distributed
in a 2D manner, that is, each process receives a subblock that is not
a block of full columns or rows. This means that the processors themselves
are, at least logically, organized in a 2D grid. Operations
then involve reductions or broadcasts inside rows or columns. For
this, a row or column of processors needs to be in a subcommunicator.

\Level 1 {Scenario: climate model}

A climate simulation code has several components, for instance corresponding
to land, air, ocean, and ice. You can imagine that each needs a different set
of equations and algorithms to simulate. You can then divide your processes,
where each subset simulates one component of the climate, occasionally communicating
with the other components.

\Level 1 {Scenario: quicksort}

The popular quicksort algorithm works by splitting the data
into two subsets that each can be sorted individually.
If you want to sort in parallel, you could implement this by making two subcommunicators,
and sorting the data on these, creating recursively more subcommunicators.

\Level 1 {Shared memory}

There is an important application of communicator splitting in the
context of one-sided communication, grouping processes by whether they
access the same shared memory area; see section~\ref{mpi-comm-split-type}.

\Level 1 {Process spawning}

Finally, newly created communicators do not always need to be subset
of the initial \indexmpishow{MPI_COMM_WORLD}.
MPI can dynamically spawn new processes (see chapter~\ref{ch:mpi-proc})
which start in a \indexmpishow{MPI_COMM_WORLD} of their own.
However, another communicator will be created that spawns the old and new worlds
so that you can communicate with the new processes.

\input chapters/mpi-commsplit

\input chapters/mpi-commgroup

\input chapters/mpi-intercomm

\newpage
\Level 0 {Review questions}

For all true/false questions, if you answer that a statement is false,
give a one-line explanation.

\begin{enumerate}

\item True or false: in each communicator, processes are numbered consecutively from zero.

\item If a process is in two communicators, it has the same rank in
both.

\item Any communicator that is not \indexmpishow{MPI_COMM_WORLD} is a strict subset of it.

\item The subcommunicators derived by \indexmpishow{MPI_Comm_split}
  are disjoint.

\item If two processes have ranks $p<q$ in some communicator,
  and they are in the same subcommunicator,
  then their ranks $p',q'$ in the subcommunicator also obey $p'<q'$.
\end{enumerate}



\index{communicator|)}

