% -*- latex -*-
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%
%%%% This text file is part of the source of
%%%% `Parallel Programming in MPI and OpenMP'
%%%% by Victor Eijkhout, copyright 2023
%%%%
%%%% otherblurb.tex : about other programming models
%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

There are many other programming systems for parallelism.
In particular threading models are a dime a dozen.

Here we discussion:
\begin{itemize}
\item \acf{CAF}: a distributed parallel mode for Fortran arrays;
  chapter~\ref{ch:caf}.
\item Kokkos and Sycl, two heterogeneous programming models that target
  both multicore and accelerator programming (and \acp{FPGA} in the case of Sycl);
  chapters \ref{ch:kokkos} and~\ref{ch:dpcpp} respectively.
\item Python multiprocessing; chapter~\ref{ch:multiprocessing}.
\end{itemize}

Comparing these to the systems already discussed above we can remark:
\begin{itemize}
\item \ac{CAF} is one of the very few alternatives to MPI where it
  comes to distributed memory programming.
  Its treatment of Cartesian arrays is more elegant; otherwise it lacks much MPI functionality.
\item Kokkos and Sycl are competitors to OpenMP offloading; chapter~\ref{ch:omp-gpu}.
  Switching between CPU and GPU modes is easier in these systems than in OpenMP.
\item The python multiprocessing toolbox is more task-base
  then the \indexterm{mpi4py} module in Python.
\end{itemize}

Among the various threading models in existence we mention
\begin{itemize}
\item \indexterm{pthreads}, which is more geared to systems programming than
  scientific computing; see \HPSCref{sec:thread-example}.
\item C++ has a native thread library; see \ISPref{sec:cpp-thread}.
  There is also the execution policy mechanism for the C++ standard library `algorithms'.
\item \indextermbus{Intel}{TBB}\index{TBB|see{Intel, TBB}} is often used as a
  lower layer for other threading implementations, such as the native C++
  parallel execution policies.
\end{itemize}

Parallelism models we do not discuss include
\begin{itemize}
\item \indexacf{CUDA} which targets \acp{GPU}.
\item \indexacf{NCCL}, which optimized MPI collectives for \acp{GPU}.
\item Chapel, a completely independent language for parallel computing;
  see \HPSCref{sec:Chapel}.
  On the topic of parallel languages, see more generally \HPSCref{sec:pgas}.
\end{itemize}
